Pleiotropy of gene regulatory variants increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2022-11-14 ms/ms02_intro.tex
MS intro

Genome-wide association studies (GWAS) allow to define the impact of genetic variants on a trait in the population.
%
The number of GWAS has rapidly increased since the first GWAS in 2010 and has result in comprehensive databases such as the NHGRI-EBI GWAS Catalog \citep{2018.Parkinson.Buniello}.
%
For instance, in 2019, NHGRI-EBI GWAS Catalog contained 5687 GWAS comprising 71673 variant-trait associations from 3567 publications \citep{2018.Parkinson.Buniello}.

However the molecular mechanism of GWAS variants is most often unclear because of three reasons \citep{2020.Trynka.CanoGamez}.
%
First, identification of the causaul variant is difficult because of linkage disequilibrium.
%
Second, GWAS do not provide hints regarding the tissues involved in the phenotypes.
%
Third, GWAS variants affect most often non-coding regions.

Expression QTLs (eQTLs) are variants that change the expression of a given gene in a tissue or cell type.
%
eQTLs share some of the same shortcommings as GWAS variants such as inability to identify the causal variant.
%
Also the distribution of eQTLs and GWAS variants in the genome have been suggested to be different because of negative selection.

Nevertheless eQTLs are still important data sets to interprete GWAS variants, because they provide gene regulatory information about the eQTL genes (eGenes) and tissues (etissues) (REF).
%
In addition colocalization anlaysis can pinpoint more precisely the causal variants that either data set independently (REF).

The large number of GWAS has allowed to discover that many genetic variants are involved in different traits \citep{2019.Posthuma.Watanabe}.
%
In this work, the analysed loci covered 60\% of the genome and 90\% of these loci contained associations across different traits \citep{2019.Posthuma.Watanabe}.
%
They also found that pleiotropic genes and variants where less tissue specific in terms of gene expression and eQTLs \citep{2019.Posthuma.Watanabe}.
%
The question of tissue specificity of eQTLs has been analyzed in different works with conflicting conclusions \citep{2021.Li.Mu.GenomeBiology.impactcelltype,2018.Vijayanand.Schmiedel.Cell.ImpactGeneticPolymorphisms,2017...Nature}.

In this work, I have taken advantage of two well formatted eQTL and GWAS data sets to develop a global colocalization analysis that includes 127 eQTL studies and 413 GWAS.
%
I have used these colocalization results to investigate molecular differences between pleiotropic variants in terms of gene regulatory properties.
%
We provide a model to mechanistically understand our results.
%
This global colocalization analysis is also useful to help interpret GWAS variants.

>>>>

% GWAS in general
Genome-wide association studies (GWAS) have allowed the association of frequent genetic variants with common traits.
%
The number of GWAS has rapidly increased since the first GWAS around 15 years ago and has resulted in comprehensive databases such as the NHGRI-EBI GWAS Catalog \citep{2007...Nature,2018.Parkinson.Buniello}.
%
For instance, in 2019, NHGRI-EBI GWAS Catalog contained 5687 GWAS comprising 71,673 variant-trait associations from 3,567 publications \citep{2018.Parkinson.Buniello}.

% Shortcomings of GWAS
However, the interpretation of the GWAS variants is complex because of three reasons \citep{2020.Trynka.CanoGamez}.
%
First, the causal variant is hidden in regions with high linkage disequilibrium (LD).
%
Second, GWAS do not inform about the genes, cell types, and tissues involved in the phenotypes.
%
Third, GWAS variants often affect non-coding regions, where the molecular cause is more difficult to define.

% Interest in studying GWAS
Expression QTLs (eQTLs) are variants that change the expression of a given gene in a cell type or tissue.
%
eQTLs share some of the same shortcomings as GWAS variants such as the inability to identify the causal variant in regions of high LD.
%
Moreover, the distribution of eQTLs and GWAS variants in the genome are different because of negative selection \citep{2022.Pritchard.Mostafavi}.
%
Nevertheless, eQTLs are still important data sets to interpret GWAS variants, because they provide gene regulatory information about the eQTL genes and cell types and tissues \citep{2021.Li.Mu}.
%
%In addition, colocalization analysis can help prioritize causal variants.

Many GWAS have allowed observing that many genetic variants are involved in different traits \citep{2019.Posthuma.Watanabe}.
%
In that work, the analyzed loci covered 60\% of the genome, and 90\% of these loci contained associations across different traits \citep{2019.Posthuma.Watanabe}.
%
They also found that pleiotropic genes and variants were less tissue-specific in terms of gene expression and eQTLs \citep{2019.Posthuma.Watanabe}.
%
However, the question of tissue specificity of eQTLs has been analyzed in different works with conflicting conclusions \citep{2021.Li.Mu,2018.Vijayanand.Schmiedel,2017gtex.nature}.
%
In addition, the mechanism of pleiotropy is not well understood yet.

In this work, I have taken advantage of two large eQTL and GWAS data sets from the EBI eQTL Catalogue and the IEU OpenGWAS, respectively,
to develop a global colocalization analysis based on 127 eQTL studies and 418 GWAS across many cell types, tissues, and phenotypes.
%
I have used these eQTL/GWAS variants to investigate mechanisms of pleiotropic gene regulatory variants.
%
These eQTL/GWAS variants might be a useful resource for the annotation and prioritization of frequent variants.
%
Therefore, I have developed a web application to query the variants.
